---
title: "CRC summary statistics"
output: html_notebook
---

# Identify relevant workspaces

This notebook is intended to illustrate combining data from multiple PRIMED 
workspaces for an analysis. We will take the example of a researcher studying
colorectal cancer who wants to use summary statistics from two studies: the
UKBB and published results from a GWAS. Both studies are available in PRIMED
open access workspaces.

- https://anvil.terra.bio/#workspaces/primed-data-cc-open/PRIMED_PAN_UKBB_GSR_CANCER
- https://anvil.terra.bio/#workspaces/primed-data-primed-cancer-open/PRIMED_CRC_PMID36539618

To fetch data from a workspace, we need two pieces of information: the workspace name and the namespace, also known as the billing project. The URL for every AnVIL workspace takes the form `<https://anvil.terra.bio/#workspaces/namespace/workspace_name>`. Therefore, we can always find this information for a workspace by navigating to it in the Workspaces menu on the AnVIL platform. Based on the URLs listed above, we can identify the workspace names and namespaces to be: 

```{r}
ukbb_workspace <- "PRIMED_PAN_UKBB_GSR_CANCER"
ukbb_namespace <- "primed-data-cc-open"
crc_workspace <- "PRIMED_CRC_PMID36539618"
crc_namespace <- "primed-data-primed-cancer-open"
```

# Retrieve data from UKBB workspace

We start with the UKBB Cancer workspace. Using the `avtable()` function from the AnVIL package, we can read in the `analysis` table to identify relevant analyses for our project. Each row of the `analysis` table corresponds to a different analysis with summary statistics in the workspace. See the PRIMED Data Model for details on all of the columns in this table. The `trait` column gives the name of the trait variable for that analysis. 

```{r}
library(AnVIL)
library(dplyr)

ukbb_analysis <- avtable("analysis", 
                    name = ukbb_workspace,
                    namespace = ukbb_namespace)

count(ukbb_analysis, trait)
```

There are four analyses of colorectal cancer, in different populations. Each row in the `analysis` table is assigned an `analysis_id` when it is imported to AnVIL, which is the primary key for this table. 

```{r}
ukbb_crc <- ukbb_analysis %>%
  filter(trait == "Colorectal cancer")

select(ukbb_crc, analysis_id, trait, population_descriptor, population_labels)
```

The `gsr_file` table contains bucket paths to the data from these analyses, one file per row. See the PRIMED Data Model for details on all of the columns in this table. The `analysis_id` links the analysis to files in the `gsr_file` table, so we use it to identify the files in the `gsr_file` table corresponding to the analysis we want (in this case, the files from the multi-population analysis).

```{r}
ukbb_gsr_file <- avtable("gsr_file", 
                    name = ukbb_workspace,
                    namespace = ukbb_namespace) %>%
  filter(analysis_id == "93f9b24b")

select(ukbb_gsr_file, analysis_id, chromosome, file_path) %>%
  head(n=2)

count(ukbb_gsr_file, analysis_id)
```

This analysis has 23 files, one file per chromosome. We use the `gsutil_cp()` function to copy the files to our current cloud environment.

```{r}
dir.create("UKBB_CRC")
gsutil_cp(ukbb_gsr_file$file_path, "UKBB_CRC")
```


# Retrieve data from GWAS workspace

The CRC GWAS workspace contains only a single analysis.

```{r}
crc_analysis <- avtable("analysis", 
                    name = crc_workspace,
                    namespace = crc_namespace)

select(crc_analysis, analysis_id, trait, population_descriptor, population_labels)
```

The `gsr_file` table contains bucket paths to the data for this analysis.

```{r}
crc_gsr_file <- avtable("gsr_file", 
                    name = crc_workspace,
                    namespace = crc_namespace)

select(crc_gsr_file, analysis_id, chromosome, file_path)
```

This analysis has only one associated file, with the results for all chromosomes. We use the `gsutil_cp()` function to copy the files to our current cloud environment.

```{r}
dir.create("GWAS_CRC")
gsutil_cp(crc_gsr_file$file_path, "GWAS_CRC")
```


# Write data tables to the workspace

To more easily keep track of the data we're working with, we can create 
`analysis` and `gsr_file` tables in this workspace. Since all data is in the
PRIMED data model, it is straightforward to concatenate tables. The AnVIL 
package has a function to import data tables to a workspace. The `entity`
argument should contain the primary key for the table.

```{r}
bind_rows(crc_analysis, ukbb_analysis) %>%
  avtable_import(entity = "analysis_id")
bind_rows(crc_gsr_file, ukbb_gsr_file) %>%
  avtable_import(entity = "gsr_file_id")
```


