---
title: "Prepare phenotype file for GWAS"
output: html_notebook
---

This notebook demonstrates how to prepare a phenotype file for use with the
[primed_genesis_gwas](https://github.com/UW-GAC/primed-genesis-gwas) workflow. 

Load R libraries:

```{r}
library(AnVIL)
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
```

Phenotype file paths are stored in the "phenotype_harmonized" data table in a 
PRIMED workspace. We use the `avtable` function to read this table, and find 
the file paths for the phenotype files we need for this analysis. In this example,
we are using a simulated LDL phenotype found in the "pilot" phenotype file.

We also need the population_descriptor phenotype file, as we will use "population"
as a covariate in the analysis. For each table, we copy the file from the google
bucket to our compute instance and read it into R.

```{r}
pheno_tables <- avtable("phenotype_harmonized")
read_phen_table <- function(table_name) {
  phen_file <- pheno_tables %>%
    filter(domain == table_name) %>%
    select(file_path) %>%
    unlist()
  gsutil_cp(phen_file, ".")
  read_tsv(basename(phen_file))
}
```

```{r}
subj <- avtable("subject")
phen <- read_phen_table("pilot")
pop <- read_phen_table("population_descriptor")
```

The population_descriptor [data model](https://github.com/UW-GAC/primed_data_models/blob/main/PRIMED_phenotype_data_model.json#L58)
allows multiple descriptors for each subject, with a delimiter. Below is a function
that extracts a single descriptor and corresponding labels from the table.

```{r}
select_pop_desc <- function(pop_desc_table, descriptor) {
  dat <- pop_desc_table %>%
    separate_longer_delim(starts_with("population"), delim="|") %>%
    mutate(across(starts_with("population"), str_trim)) %>%
    pivot_wider(names_from=population_descriptor, values_from=population_label) %>%
    select(subject_id, all_of(descriptor))
}
```

We select the outcome and covariates we plan to use in the GWAS. The GENESIS workflow
expects a column called 'sex' with values 'M' and 'F', so these values must be 
mapped from the PRIMED data model.

```{r}
phen_gwas <- subj %>%
  mutate(sex=c("Male"="M", "Female"="F")[reported_sex]) %>%
  inner_join(phen) %>%
  select(subject_id, sex, age_at_observation, ldl) %>%
  inner_join(select_pop_desc(pop, "population"))
```

In the PRIMED data model, the sample table links subject identifiers (subject_id)
to the sample identifiers in the genotype files (sample_id). The GENESIS workflow
requires sample_id to match phenotypes to genotypes.

```{r}
sample <- avtable("sample")
```

```{r}
phen_gwas_samples <- sample %>%
  select(sample_id, subject_id) %>%
  inner_join(phen_gwas)
```

The GENESIS workflow expects a phenotype file in CSV format. We save the
phenotype data table as CSV and copy it to the workspace bucket.

```{r}
outfile <- "phenotypes_ldl_gwas.csv"
bucket_path <- file.path(avbucket(), "GWAS", outfile)
write_csv(phen_gwas_samples, outfile)
gsutil_cp(outfile, bucket_path)
```
